---
title: "Stat 240 Final Project"
author: "Ami, Andrew, Indu"
output: html_document
date: "2024-12-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 
Poverty and homelessness have always been prevailing issues in the U.S. and in recent years have become even more prominent issues. Housing prices have been rising and many Americans find it hard to afford housing. Many people ask the question is there anything being done to fix this issue, and many claim that there isn’t enough being done. To answer this question, we will use data from NYC Open Data to compare the number of low-income housing units that are produced to that of middle-income households. In this report we will show that there is a statistically significant difference in the proportions of low-income households and middle-income households that have been produced in New York City since 2014. Meaning that there are significant efforts being made to help resolve the housing crisis especially for low-income individuals.  

## Background 
The dataset we decided to use is published on the NYC Open Data website. The Department of Housing Preservation Development generates reports on projects, buildings, and units that began development after January 1, 2014. Each row of data refers to a different housing project in New York City. The key variables are very low-income units’ middle-income units and project completion date. It is these metrics that we conducted our analysis on. 

## Analysis
```{r echo = FALSE}
library(tidyverse)

# Read the CSV
housing_data <- read_csv("Affordable_Housing_Production_by_Project_20241202.csv")

# Data Preprocessing
processed_data <- housing_data %>%
  # Remove rows with NA in key columns
  drop_na(`Project Start Date`, `Very Low Income Units`, `Middle Income Units`, `Total Units`) %>%
  # Parse dates using specific format
  mutate(
    project_start_date = as.Date(`Project Start Date`, format = "%m/%d/%Y"),
    project_start_year = year(project_start_date)
  ) %>%
  # Filter for 2014-2024 range
  filter(
    project_start_year >= 2014 & 
    project_start_year <= 2024
  )

# 1. Bar Chart of Housing Units
bar_chart <- processed_data %>%
  group_by(project_start_year) %>%
  summarise(
    very_low_income = sum(`Very Low Income Units`, na.rm = TRUE),
    middle_income = sum(`Middle Income Units`, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(very_low_income, middle_income), 
               names_to = "income_category", 
               values_to = "units") %>%
  ggplot(aes(x = project_start_year, y = units, fill = income_category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Affordable Housing Units by Income Category",
    x = "Year",
    y = "Number of Units",
    fill = "Income Category"
  ) +
  theme_minimal()

# 2. Pie Chart of Unit Proportions
pie_chart_data <- processed_data %>%
  summarise(
    very_low_income = sum(`Very Low Income Units`, na.rm = TRUE),
    middle_income = sum(`Middle Income Units`, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "category", values_to = "units")

pie_chart <- pie_chart_data %>%
  ggplot(aes(x = "", y = units, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(
    title = "Proportion of Housing Units by Income Category",
    fill = "Income Category"
  ) +
  theme_void()

# 3. Time Series Line Graph
time_series <- processed_data %>%
  group_by(project_start_year) %>%
  summarise(
    very_low_income = sum(`Very Low Income Units`, na.rm = TRUE),
    middle_income = sum(`Middle Income Units`, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(very_low_income, middle_income), 
               names_to = "income_category", 
               values_to = "units") %>%
  ggplot(aes(x = project_start_year, y = units, color = income_category, group = income_category)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Housing Units Over Time by Income Category",
    x = "Year",
    y = "Number of Units",
    color = "Income Category"
  ) +
  theme_minimal()

# Statistical Analysis
total_very_low <- sum(processed_data$`Very Low Income Units`, na.rm = TRUE)
total_middle <- sum(processed_data$`Middle Income Units`, na.rm = TRUE)
total_units <- sum(processed_data$`Total Units`, na.rm = TRUE)

prop_very_low <- total_very_low / total_units
prop_middle <- total_middle / total_units

# Two-proportion z-test (manual calculation)
pooled_prop <- (total_very_low + total_middle) / (2 * total_units)
se <- sqrt(pooled_prop * (1 - pooled_prop) * (2 / total_units))
z_stat <- (prop_very_low - prop_middle) / se
p_value <- 2 * (1 - pnorm(abs(z_stat)))

# Print results
cat("Two-Proportion Z-Test Results:\n")
cat("Z-statistic:", z_stat, "\n")
cat("P-value:", p_value, "\n")
cat("Proportion of Very Low Income Units:", prop_very_low, "\n")
cat("Proportion of Middle Income Units:", prop_middle, "\n")

# Optional: Print all 3 plots
print(bar_chart)
print(pie_chart)
print(time_series)


```

```{r}

processed_data <- housing_data %>%
  mutate(
    project_start_year = as.numeric(str_extract(`Project Start Date`, "\\d{4}")),
    project_start_month = as.numeric(str_extract(`Project Start Date`, "^\\d+"))
  ) %>%
  filter(project_start_year >= 2014 & project_start_year <= 2024)

# Data Preprocessing
processed_data <- housing_data %>%
  drop_na(`Project Start Date`, `Very Low Income Units`, `Middle Income Units`, `Total Units`) %>%
  mutate(
    project_start_date = as.Date(`Project Start Date`, format = "%m/%d/%Y"),
    project_start_year = year(project_start_date),
    project_start_month = month(project_start_date, label = TRUE)
  ) %>%
  filter(project_start_year >= 2014 & project_start_year <= 2024)

# Yearwise Analysis
yearwise_analysis <- processed_data %>%
  group_by(project_start_year) %>%
  summarise(
    very_low_units = sum(`Very Low Income Units`),
    middle_units = sum(`Middle Income Units`),
    total_units = sum(`Total Units`),
    very_low_prop = very_low_units / total_units,
    middle_prop = middle_units / total_units
  )

# Monthwise Analysis
monthwise_analysis <- processed_data %>%
  group_by(project_start_month) %>%
  summarise(
    very_low_units = sum(`Very Low Income Units`),
    middle_units = sum(`Middle Income Units`),
    total_units = sum(`Total Units`),
    very_low_prop = very_low_units / total_units,
    middle_prop = middle_units / total_units
  )

# Yearwise Visualization
yearwise_plot <- yearwise_analysis %>%
  pivot_longer(cols = c(very_low_units, middle_units), 
               names_to = "unit_type", 
               values_to = "units") %>%
  ggplot(aes(x = project_start_year, y = units, fill = unit_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Yearly Housing Units by Income Category",
       x = "Year", y = "Number of Units") +
  theme_minimal()

# Monthwise Visualization
monthwise_plot <- monthwise_analysis %>%
  pivot_longer(cols = c(very_low_units, middle_units), 
               names_to = "unit_type", 
               values_to = "units") %>%
  ggplot(aes(x = project_start_month, y = units, fill = unit_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Monthly Housing Units by Income Category",
       x = "Month", y = "Number of Units") +
  theme_minimal()

# Print analyses and plots
print(yearwise_analysis)
print(monthwise_analysis)
print(yearwise_plot)
print(monthwise_plot)
# Yearly Z-Test
yearly_data <- housing_data %>%
  mutate(year = as.numeric(str_extract(`Project Start Date`, "\\d{4}"))) %>%
  filter(year >= 2014 & year <= 2024) %>%
  group_by(year) %>%
  summarise(
    very_low_units = sum(`Very Low Income Units`, na.rm = TRUE),
    middle_units = sum(`Middle Income Units`, na.rm = TRUE),
    total_units = sum(`Total Units`, na.rm = TRUE)
  ) %>%
  mutate(
    very_low_prop = very_low_units / total_units,
    middle_prop = middle_units / total_units,
    pooled_prop = (very_low_units + middle_units) / (2 * total_units),
    se = sqrt(pooled_prop * (1 - pooled_prop) * (2 / total_units)),
    z_stat = (very_low_prop - middle_prop) / se,
    p_value = 2 * (1 - pnorm(abs(z_stat)))
  )

# Monthly Z-Test
monthly_data <- housing_data %>%
  mutate(month = as.numeric(str_extract(`Project Start Date`, "^\\d+"))) %>%
  group_by(month) %>%
  summarise(
    very_low_units = sum(`Very Low Income Units`, na.rm = TRUE),
    middle_units = sum(`Middle Income Units`, na.rm = TRUE),
    total_units = sum(`Total Units`, na.rm = TRUE)
  ) %>%
  mutate(
    very_low_prop = very_low_units / total_units,
    middle_prop = middle_units / total_units,
    pooled_prop = (very_low_units + middle_units) / (2 * total_units),
    se = sqrt(pooled_prop * (1 - pooled_prop) * (2 / total_units)),
    z_stat = (very_low_prop - middle_prop) / se,
    p_value = 2 * (1 - pnorm(abs(z_stat)))
  )

print(yearly_data[, c("year", "very_low_prop", "middle_prop", "z_stat", "p_value")])
print(monthly_data[, c("month", "very_low_prop", "middle_prop", "z_stat", "p_value")])



```

## Discussion 
### Statistical Significance
The extremely small p-values (0 in most years) indicate that the differences between very-low-income and middle-income housing proportions are not random - they represent real, deliberate policy choices in housing development.  

### Shortcomings
No information about total housing demand for each income category
Missing data on actual occupancy rates versus just unit construction

### Additional Questions to consider
How many people from each income category actually obtained and retained housing?
How do these numbers compare to other major cities?
